Highly Available Transactions: Virtues and Limitations

### 2013-05-01 writeup/hat-analysis.tex
pass, plus spelling and grammar

HAT systems provide transactions with transactional availability and
either high availability or sticky high availability. They offer
latency and availability benefits over traditional distributed
databases, yet they cannot achieve all possible semantics. In this
section, we delineate ACID, distributed consistency, and session
consistency levels which can be achieved with high availability
(transactional atomicity, variants of Repeatable Read isolation, and
many session guarantees), those with sticky high availability (read
your writes, PRAM and causal consistency), and properties that cannot
be provided in a HAT system (those preventing Lost Update and Write
Skew, or with recency).  We present a full summary of these results in
Section~\ref{sec:hat-summary}.

@@

To begin, we present well-known semantics that can be achieved in HAT
systems. In this section, our goal is feasibility, not performance. As
a result, we offer proof-of-concept highly available algorithms, but
our goal is \textit{not} to provide optimal or even efficient
implementations: the challenge is to provide algorithms that provide
high availability. However, we briefly study performance implications
in Section~\ref{sec:evaluation}.

\subsubsection{ACID Isolation Guarantees}

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}. In this model, writes to each object are totally
ordered, corresponding to the order in which they are installed in the
database. In a distributed database, different replicas may receive
writes to their local copies of data at different times, but the order
in which replicas overwrite versions should be consistent with the total order for
each item. \textit{PL-1} requires that writes to different objects are
ordered consistently across transactions, prohibiting Adya's
phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If we
create a graph of transactions with edges from one transaction to
another if the former overwrites the latter's write to the same
object, then, under Read Uncommitted, there should be no cycles in
this graph. Consider the following example:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\end{align*}
In this example, under Read Uncommitted, it is impossible for the
database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
achieved via timestamping each update in a transaction with the same
logical timestamp (which is unique across transactions; e.g.,
combining a client ID and logical clock) and applying a ``last writer
wins'' conflict reconciliation policy at each replica. Later
properties will strengthen Read Uncommitted.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs. Centralized
implementations differ, with some based on long-duration exclusive
locks and short-duration read locks~\cite{gray-isolation} and others
based on multiple versions. The implementations often provide recency
and monotonicity properties beyond the simple meaning from the name,
which is what is expressed in the implementation-agnostic definition:
under Read Committed, transactions should not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should never read
$a=3$:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-1em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of data, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. This implementation does not provide recency or
monotonicity guarantees but satisfies the implementation-agnostic
definition.

Several different properties have been labeled \textbf{Repeatable
  Read} Isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures directly the spirit of the term: if a
transaction reads the same data twice, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation (I-CI)}, and if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s
$P3/A3$~\cite{ansicritique}), we have the stronger property of
\textbf{Predicate Cut-Isolation}. In the example below, under both
levels of cut isolation, $T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having clients cache their reads such that the value that they read
never changes unless they overwrite it themselves. The cache can be
cleared at the end of each transaction and can alternatively be
accomplished on (sticky) replicas via multi-versioning. Predicate Cut
Isolation is also achievable in HAT systems via similar caching or
multi-versioning that tracks entire logical ranges of predicates
in addition to item-based reads.

\subsubsection{ACID Atomicity Guarantees}

\textbf{Transactional atomicity (TA)} is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property, TA
restricts transactions' ability to view the effects of partially
completed transactions. Under TA, once some of the effects of a
transaction $T_i$ are observed by another transaction $T_j$,
thereafter, all effects of $T_i$ are observed by $T_j$. Together with
item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perplexingly, discussions of TA are absent from existing treatment of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), TA is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore}. In contrast,
in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to ensure that, regardless of replica
selection, all of a client's accesses will read a sufficiently
up-to-date/``consistent'' set of data items. However, replicas do not
need to agree on when to choose to reveal new sets of values clients,
which would require consensus. Instead, TA only requires the
equivalent of reliable broadcast (with some additional client
metadata) and is achievable in HAT systems.

TA can be achieved via multi-versioning. If clients attach vector
clocks to every write (incrementing their own position for each
transaction) and replicas store all writes via multi-versioning, then
clients can safely determine which sets of data items to read; this
approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
bolt-on causal consistency~\cite{bolton}. These systems also provide
causal consistency and are subsequently sticky available (each is
implemented via client-side caching). However, TA is achievable
without stickiness via an alternate algorithm: if servers wait to
reveal writes until they are present on all replicas, clients do not
need to be sticky.

We sketch a non-sticky TA algorithm and provide greater detail in our
technical report~\cite{hat-tr}. Each replica waits to reveal write $w$
from transaction $T$ until all of the writes from $T$ are present on
all respective replicas (are \textit{pending stable}). Once they are,
replicas asynchronously reveal their respective writes to readers. To
prevent clients from reading incomplete portions of a transaction's
writes as they are being revealed (e.g., a client reads a data item
that from a replica that has detected pending stability, then attempts
to read another item from another replica which has not yet detected
pending stability), clients include additional metadata with each
write: a list of items in the transaction along with the transaction's
timestamp (from Read Uncommitted). When a client reads a write, the
write's list of items and timestamp form a lower bound on the versions
that the client should read for other items. When a client performs
subsequent reads, it includes a timestamp with its request that
corresponds current lower bound on reads from the respective item
Replicas respond with either a write matching the timestamp or a write
with a higher timestamp that is pending stable. Clients discard any
metadata upon transaction commit, and servers keep two sets of writes
for each data item: the write with the highest timestamp that is
pending stable and a set of writes that are not yet pending stable.

This non-sticky implementation is entirely masterless and neither
reads nor writes block for coordination. Moreover, there are several
possible optimizations that, for brevity, we do not describe
here. However, one key property is that we do not, in general, dictate
when writes become visible. Accordingly, the algorithm is highly
available. We can optimize write visibility by sticking clients are
with disjoint groups of servers: once all of a transaction's writes
are pending stable within a group of servers, writes can be made
visible (as opposed to once they are pending stable with respect to
all servers). For example, in a multi-datacenter setting, all clients
and servers within a datacenter may form a sticky group.

@@

As is typical in the database literature (with few
exceptions~\cite{daudjee-session}), our models have not yet considered
interactions \textit{across} transactions, other than that there
exists some arbitrary ordering on them (via Read Uncommitted). In the
distributed systems literature, many useful \textit{safety} guarantees
span multiple, non-transactional operations. In particular,
\textit{session guarantees} are used to describe guarantees across
transactions within a given \textit{session}, ``an abstraction for the
sequence of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

\vspace{.5em}\noindent\textbf{{Monotonic reads}} requires that, within
a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress according to a total
order. The ordering of versions read should respect any ordering on
transactions.

@@

The above guarantees can be achieved by forcing servers to wait to
reveal new writes until each write's respective dependencies are
fulfilled on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written. This \textit{is} highly available as a client
will never block due to inability to find a server with a sufficiently
up-to-date version of a data item. However, it does not imply that
transactions will read their own writes or, in the presence of
partitions, make forward progress through the version history. The
problem is that, if a server becomes partitioned, under the highly
available model, we must handle the possibility that an unfortunate
client will be forced to issue her next requests against the
partitioned server.

The solution to this conundrum is to give up high availability in
favor of sticky availability. Sticky availability permits three
additional guarantees, which we first define and then prove are
unachievable in a generic highly available system:

\vspace{.5em}\noindent\textbf{{Read your writes}} requires
that whenever a client reads a given data item after updating it, the
read returns the updated value (or a value with a higher ID).

\vspace{.5em}\noindent\textbf{{PRAM}} (Pipelined Random Access
Memory) provides the illusion of serializing each session's operations
and is the combination of monotonic reads, monotonic writes, and read
your writes~\cite{herlihy-art}.

\vspace{.5em}\noindent\textbf{{Causal
    consistency}}~\cite{causalmemory} is the combination of all of the
session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
writes-follow-reads) and is also referred to by Adya as PL-2L
isolation~\cite{adya}).\vspace{.5em}

Read your writes is not achievable in a highly available
system. Consider a client that executes the following two transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)
\\T_2 &: r_x(a)\vspace{-1em}
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client subsequently executes $T_2$ against the same
(partitioned) server, then it will be able to read its
writes. However, if the network topology shifts and the client can
only contact a different server that is partitioned from the server
that executed $T_1$, then the client will be unable to read its own
writes and the system will have to either stall indefinitely to allow
the client to read her writes (violating transactional availability)
or will have to sacrifice read your writes guarantees. However, if the
client remains sticky with the server that executed $T_1$, then we can
disallow this scenario. Accordingly, read your writes, and, by proxy,
causal consistency and PRAM require stickiness. Read your writes is
provided by default in a sticky system. Causality and PRAM guarantees
can be accomplished with well-known variants~\cite{causalmemory,
  bolton, cops, sessionguarantees, swift} of the prior session
guarantee algorithms we presented earlier.

@@

In this section, we briefly discuss additional noteworthy guarantees
achievable by HAT systems.

\vspace{0.5em}
\noindent{\textbf{Consistency}} A HAT system can make limited
application-level consistency guarantees. It can often execute
commutative and logically monotonic~\cite{calm} operations without the
risk of invalidating application-level integrity constraints. We do
not attempt to sketch the entire space of \textit{application-level}
consistency properties that are achievable in HAT systems (see
Section~\ref{sec:relatedwork}) but we specifically evaluate TPC-C
transaction semantics under HAT consistency guarantees in
Section~\ref{sec:evaluation}.

@@

\vspace{.5em}\noindent{\textbf{Convergence}} To require that the
system propagates writes between replicas, we can require convergence,
or eventual consistency for each data item: in the absence of new
mutations to a data item, in the absence of partitions, all servers
should eventually agree on the value for each item. This is typically
accomplished by any number of anti-entropy protocols, which
periodically update neighboring servers with the latest value for each
data item~\cite{antientropy}. Establishing a final value is related to
determining a total order on transaction updates, as in Read
Uncommitted.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and one-copy serializability---inherently requires unavailability.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small\vspace{-1em}
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-1em}
\end{align*}
Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
database state could not have arisen serial execution of $T_1$ and
$T_2$.\footnote{In this example, we assume that, as is standard in
  modern databases, replicas accept values as they are written (i.e.,
  register semantics). This particular example could be made
  serializable via the use of commutative updates
  (Section~\ref{sec:evaluation}) but the problem persists in the
  general case.}  To prevent this from happening, either $T_1$ or
$T_2$ should not have committed. Each client's respective server might
try to detect that another write occurred, but this requires knowing
the version of the latest write to $x$. In our example, this reduces
to a requirement for linearizability, which is, via Gilbert and
Lynch's proof of the CAP Theorem, provably unachievable with high
availability~\cite{gilbert-cap}.

@@

Distributed data storage systems often make various recency guarantees
on reads of data items. As we have discussed, one of the most famous
is linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g.,
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force an available system to violate any recency bound, so recency
bounds are not enforceable by HAT systems.

@@

To the best of our knowledge, this is the first unification of
database ACID transactions, distributed consistency, and session
guarantee models. Interestingly, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 96 possible HAT
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonomy is not exhaustive, but we believe it
lends substantial clarity into the relationships between a large
subset of the prominent ACID and distributed consistency
models. Additional read/write transaction semantics that we have
omitted should be easily classifiable based on the available
primitives and HAT-incompatible anomaly prevention we have already
discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (I-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \draw [->] (ru) -- (rc);
 \draw [->] (rc) -- (ta);
 \draw [->] (ici) -- (pci);

>>>>

HAT systems provide transactions with transactional availability and
either high availability or sticky high availability. They offer
latency and availability benefits over traditional distributed
databases, yet they cannot achieve all possible semantics. In this
section, we delineate ACID, distributed replica consistency, and
session consistency levels which can be achieved with high
availability (transactional atomicity, variants of Repeatable Read and
traditionally weaker isolation, and many session guarantees), those
with sticky high availability (read your writes, PRAM and causal
consistency), and properties that cannot be provided in a HAT system
(those preventing Lost Update and Write Skew or guaranteeing recency).
We present a full summary of these results in
Section~\ref{sec:hat-summary}.

@@

To begin, we present well-known semantics that can be achieved in HAT
systems. In this section, our primary goal is feasibility, not
performance. As a result, we offer proof-of-concept highly available
algorithms that are not necessarily optimal or even efficient: the
challenge is to prove the existence of algorithms that provide high
availability. However, we briefly study a subset of their performance
implications in Section~\ref{sec:evaluation}.

\subsubsection{ACID Isolation Guarantees}
\label{sec:isolation}

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}. In this model, writes to each object are totally
ordered, corresponding to the order in which they are installed in the
database. In a distributed database, different replicas may receive
writes to their local copies of data at different times but should
handle concurrent updates (i.e., overwrites) in accordance with the
total order for each item. \textit{PL-1} requires that writes to
different objects be ordered consistently across transactions,
prohibiting Adya's phenomenon $G0$ (also called ``Dirty
Writes''~\cite{ansicritique}). If we build a graph of transactions
with edges from one transaction to another if the former overwrites
the latter's write to the same object, then, under Read Uncommitted,
the graph should not contain cycles~\cite{adya}. Consider the
following example:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\end{align*}
In this example, under Read Uncommitted, it is impossible for the
database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
achieved via timestamping each update in a transaction with the same
timestamp (unique across transactions; e.g., combining a client's ID
with a sequence number) and applying a ``last writer wins'' conflict
reconciliation policy at each replica. Later properties will
strengthen Read Uncommitted.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs (as in
Section~\ref{sec:modernacid}). Centralized implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. These implementations often provide recency and monotonicity
properties beyond what is implied by the name ``Read Committed'' and
what is captured by the implementation-agnostic definition: under
Read Committed, transactions should not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read
$a=3$:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-1em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of data, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. Unlike a lock-based implementation, this implementation
does not provide recency or monotonicity guarantees but it satisfies
the implementation-agnostic definition.

Several different properties have been labeled \textbf{Repeatable
  Read} isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures the spirit of the term: if a
transaction reads the same data twice, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation}, and, if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
the stronger property of \textbf{Predicate Cut-Isolation}. In the
example below, under both levels of cut isolation, $T_3$ must read
$a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having transactions cache their reads at the client such that the
values that they read for each item never changes unless they
overwrite it themselves. The cache can be cleared at the end of each
transaction and can alternatively be accomplished on (sticky) replicas
via multi-versioning. Predicate Cut Isolation is also achievable in
HAT systems via similar caching or multi-versioning that track entire
logical ranges of predicates in addition to item-based reads.

\subsubsection{ACID Atomicity Guarantees}
\label{sec:ta}

\textbf{Transactional atomicity (TA)} is core to ACID
guarantees. Although, at least by the ACID acronym, it is not an
``isolation'' property, TA restricts transactions' ability to view the
effects of partially completed transactions. Under TA, once some of
the effects of a transaction $T_i$ are observed by another transaction
$T_j$, thereafter, all effects of $T_i$ are observed by
$T_j$. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A). In the example below, under TA,
because $T_2$ has read $T_1$'s write to $y$, $T_2$ must observe
$b=c=1$ (or later versions for each key):
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$x$. Notably, TA requires disallows reading intermediate writes
(Adya's $G1b$): observing all effects of a transaction implicitly
requires observing the final (committed) effects of the transaction as
well.

Perplexingly, discussions of TA are absent from existing treatments of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), TA is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore}. In contrast,
in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to ensure that, regardless of replica
selection, all of a client's accesses will read a sufficiently
up-to-date/``consistent'' set of data items. However, replicas do not
need to agree on when to choose to reveal sets of values to clients, which would require consensus. Instead, as we see below, TA
only requires the equivalent of reliable broadcast (with some
additional client metadata) and is achievable in HAT systems.

% TA can be achieved via multi-versioning. If clients attach vector
% clocks to every write (incrementing their own position for each
% transaction) and replicas store all writes via multi-versioning, then
% clients can safely determine which sets of data items to read; this
% approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
% bolt-on causal consistency~\cite{bolton}. These systems also provide
% causal consistency and are subsequently sticky available (each is
% implemented via client-side caching). However, TA is achievable
% without stickiness via an alternate algorithm: if servers wait to
% reveal writes until they are present on all replicas, clients do not
% need to be sticky.

We sketch a highly available TA algorithm and provide greater detail
in our technical report~\cite{hat-tr}. We begin with our Read
Committed algorithm, but replicas wait to reveal new writes to readers
until all of the replicas for the final writes in the transaction have
received their respective writes (are \textit{pending
  stable}). Accordingly, replicas reveal writes asynchronously;
different replicas may detect pending stability at different times. To
prevent clients from reading incomplete portions of a transaction's
writes as they are being revealed (e.g., a client reads a data item
that from a replica that has detected pending stability, then attempts
to read another item from another replica that has not yet detected
pending stability), clients include additional metadata with each
write: a single timestamp for all writes in the transaction (e.g., as
in Read Uncommitted) and a list of items written to in the
transaction. When a client reads a write, the write's timestamp and
list of items form a lower bound on the versions that the client
should read for the other items. When a client reads any data item, it
attaches a timestamp to its request representing the current lower
bound for that item. Replicas use this timestamp to respond with
either a write matching the timestamp or a pending stable write with a
higher timestamp. Clients may discard any metadata upon transaction
commit, and servers keep two sets of writes for each data item: the
write with the highest timestamp that is pending stable and a set of
writes that are not yet pending stable.

This implementation is entirely master-less and operations never block
due to replica coordination. Moreover, there are several possible
optimizations that, for brevity, we do not describe here. However, one
key property is that we do not dictate when writes become visible;
accordingly, the algorithm is highly available. We can optimize write
visibility via sticky availability by sticking clients are with
disjoint groups of servers: once all of a transaction's writes are
pending stable within a group of servers, writes can be made visible
(as opposed to once they are pending stable with respect to all
servers). For example, in a multi-datacenter setting, all clients and
servers within a datacenter may form a sticky group.

@@

As is typical in the database literature (with few
exceptions~\cite{daudjee-session}), our models have not yet considered
real-time or client-centric ordering \textit{across} transactions. In
the distributed systems literature~\cite{sessionguarantees, vogels-defs}, many
useful \textit{safety} guarantees span multiple, non-transactional
operations. In particular, \textit{session guarantees} are used to
describe guarantees across transactions within a given
\textit{session}, ``an abstraction for the sequence of...operations
performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

@@

\vspace{.5em}\noindent\textbf{{Monotonic reads}} requires that, within
a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress according to a total
order (e.g., the order from Read Uncommitted).

@@

The above guarantees can be achieved by forcing servers to wait to
reveal new writes until each write's respective dependencies are
visible on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written. This is highly available as a client will never block due to
inability to find a server with a sufficiently up-to-date version of a
data item. However, it does not imply that transactions will read
their own writes or, in the presence of partitions, make forward
progress through the version history. The problem is that under
non-sticky high availability, a system must handle the possibility
that, under a partition, an unfortunate client will be forced to issue
its next requests against a partitioned, out-of-date server.

The solution to this conundrum is to give up high availability and
settle for sticky availability. Sticky availability permits three
additional guarantees, which we first define and then prove are
unachievable in a generic highly available system:

\vspace{.5em}\noindent\textbf{{Read your writes}} requires that
whenever a client reads a given data item after updating it, the read
returns the updated value (or a value that overwrote the previously
written value).

\vspace{.5em}\noindent\textbf{{PRAM}} (Pipelined Random Access Memory)
provides the illusion of serializing each of the operations (both
reads and writes) within each session and is the combination of
monotonic reads, monotonic writes, and read your
writes~\cite{herlihy-art}.

\vspace{.5em}\noindent\textbf{{Causal
    consistency}}~\cite{causalmemory} is the combination of all of the
session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
writes-follow-reads) and is also referred to by Adya as \textit{PL-2L}
isolation~\cite{adya}).\vspace{.5em}

Read your writes is not achievable in a highly available
system. Consider a client that executes the following two transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)
\\T_2 &: r_x(a)\vspace{-1em}
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, then, for transactional availability,
the server must allow $T_1$ to commit. If the same client subsequently
executes $T_2$ against the same (partitioned) server in the same
session, then it will be able to read its writes. However, if the
network topology changes and the client can only contact a different
server that is partitioned from the server that executed $T_1$, then
the system will have to either stall indefinitely to allow the client
to read her writes (violating transactional availability) or will have
to sacrifice read your writes guarantees. However, if the client
remains sticky with the server that executed $T_1$, then we can
disallow this scenario. Accordingly, read your writes, and, by proxy,
causal consistency and PRAM require stickiness. Read your writes is
provided by default in a sticky system. Causality and PRAM guarantees
can be accomplished with well-known variants~\cite{causalmemory,
  bolton, cops, sessionguarantees, swift} of the prior session
guarantee algorithms we presented earlier.

@@

In this section, we briefly discuss two additional kinds of guarantees
that are achievable in HAT systems.

\vspace{0.5em}
\noindent{\textbf{Consistency}} A HAT system can make limited
application-level consistency guarantees. It can often execute
commutative and logically monotonic~\cite{calm} operations without the
risk of invalidating application-level integrity constraints and can
maintain limited criteria like foreign key constraints (via TA). We do
not attempt to describe the entire space of application-level
consistency properties that are achievable (see
Section~\ref{sec:relatedwork}) but we specifically evaluate TPC-C
transaction semantics with HAT guarantees in
Section~\ref{sec:evaluation}.

@@

\vspace{.5em}\noindent{\textbf{Convergence}} To require that the
system propagates writes between replicas, we can require convergence,
or \textit{eventual consistency}: in the absence of new mutations to a
data item, in the absence of partitions, all servers should eventually
agree on the value for each item~\cite{cac, vogels-defs}. This is
typically accomplished by any number of anti-entropy protocols, which
periodically update neighboring servers with the latest value for each
data item~\cite{antientropy}. Establishing a final convergent value is
related to determining a total order on transaction updates to each
item, as in Read Uncommitted.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and one-copy serializability---inherently requires foregoing high
availability guarantees.

@@

It is impossible to prevent Lost Update in a highly available
environment. Consider two clients who submit the following $T_1$ and
$T_2$ on opposite sides of a network partition:
\begin{align*}
\small\vspace{-1em}
T_1 &: r_x(100)~w_x(100+20=120)
\\T_2 &: r_x(100)~w_x(100+30=130)\vspace{-1em}
\end{align*}
Regardless of whether $x=120$ or $x=130$ is chosen by a replica, the
database state could not have arisen from a serial execution of $T_1$ and
$T_2$.\footnote{In this example, we assume that, as is standard in
  modern databases, replicas accept values as they are written (i.e.,
  register semantics). This particular example could be made
  serializable via the use of commutative updates
  (Section~\ref{sec:evaluation}) but the problem persists in the
  general case.}  To prevent this, either $T_1$ or
$T_2$ should not have committed. Each client's respective server might
try to detect that another write occurred, but this requires knowing
the version of the latest write to $x$. In our example, this reduces
to a requirement for linearizability, which is, via Gilbert and
Lynch's proof of the CAP Theorem, provably at odds with high
availability~\cite{gilbert-cap}.

@@

Distributed data storage systems often make various recency guarantees
on reads of data items.  Unfortunately, an indefinitely long partition
can force an available system to violate any recency bound, so recency
bounds are not enforceable by HAT systems~\cite{gilbert-cap}. One of
the most famous of these guarantees is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g.,
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. None of these guarantees is HAT-compliant.

@@

To the best of our knowledge, this is the first unification of
database ACID transactions, distributed consistency, and session
guarantee models. Interestingly, strong one-copy serializability
entails all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 144 possible HAT
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonomy is not exhaustive
(Section~\ref{sec:conclusion}), but we believe it lends substantial
clarity to the relationships between a large subset of the prominent
ACID and distributed consistency models. Additional read/write
transaction semantics that we have omitted should be classifiable
based on the available primitives and HAT-incompatible anomaly
prevention we have already discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Read Uncommitted (RU), Read Committed (RC), Transactional
Atomicity (TA), Item Cut Isolation (I-CI), Predicate Cut Isolation
(P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
Writes (MW)\\\hline Sticky HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \draw [->] (ru) -- (rc);
 \draw [->] (ici) -- (pci);
